Estimating duration in the presence of misclassification

### 2024-03-09 introduction.tex
restructure intro

The most famous reproduction number is the \emph{basic reproduction number}, $\R$.
$\R$ is the mean number of secondary infections generated by a single infection in a \emph{fully susceptible} population with \emph{standard behaviour patterns}.
A fully susceptible population means none of its members have any immunity to the disease.
Note that this definition is a constant for a given pathogen and population.
In the early stages of an epidemic caused by a novel pathogen, the number of individuals with immunity is negligible, and hence $\R$ describes this part of the epidemic well.
$\R$ has a threshold value of 1: if $\R < 1$ then the epidemic will die out.

@@

However, incidence and transmission are almost never directly observed.
Therefore, they need to be estimated from other data sources.

\emph{Prevalence} is the proportion of the population with the disease at a given time.
Whether an individual is prevalent (has the disease) is normally measurable.
Therefore, the prevalence of the population can be estimated with a prevalence survey.

See \cref{E-inc-prev:sec:definitions} for a formal definition of incidence and prevalence.
Prevalence and incidence are strongly related, with the relationship determined by how long the disease lasts (see \cref{E-inc-prev}).

@@

\section{Data sources} \label{intro:sec:data-sources}

A wide variety of designed studies, administrative data, and other data sources have been used for pandemic and epidemic surveillance.
In the context of the SARS-CoV-2 pandemic in England, the most prominent were: community testing (known as pillar 2 testing), the number of severe events (\eg hospital admissions or deaths), seroprevalence testing (seroprevalence is the proportion of people with antibodies to a disease, a marker of prior infection), and prevalence surveys.
Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} for further details.

Community testing is broadly available, but does not test a random sample of the population.
The number of tests conducted varies over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
For large parts of the pandemic, community testing was widely available in the UK.
Therefore, there are many test results, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.

Seroprevalence testing provides information on a marker of recent prior infection.
Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.
Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) who are not representative of the general population.

Severe events counts are generally a reliable measure of the burden of the disease.
The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context\todo{ref severe events criteria}.
Therefore, the data is subject to fewer biases than community testing.
However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe event counts are insensitive to changes in transmission only occurring in younger individuals.

@@

The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
In particular, I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal study, testing a large number of indidividuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
Early in the pandemic, a methodology was developed for using CIS data to estimate prevalence (see \cref{E-intro:sec:cis}).

@@

The primary aim of this thesis is to estimate incidence and transmission from the CIS data.

\section{Estimating incidence and transmission} \label{intro:sec:estimating-incidence}

As noted above, incidence and transmission are extremely important quantities for informing the response to a pandemic but cannot be observed directly.
Therefore, they must be inferred from observable quantities.

Many methods for estimating incidence and transmission were developed in the context of the HIV/AIDS pandemic.
% HIV is transmitted sexually or through blood and causes the disease AIDS.
This ongoing pandemic has killed 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
The large burden of the disease has motivated many advances in infectious disease epidemiology and biostatistics, including several which form the methodological background of this thesis.

\subsection{Backcalculation} \label{intro:sec:backcalc}

\emph{Backcalculation} is a statistical procedure for estimating incidence based on a measurement of prevalence.
It was proposed by \textcite{brookmeyerMethod} in the context of HIV/AIDS.
In this context, backcalculation estimates the incidence of HIV infection from data on the number of AIDS diagnoses, and the time between infection and AIDS diagnosis~\autocite{brookmeyerBackcalculation,brookmeyerMeasuring}.
Two processes needed to be considered: one generating infections and one relating infections to prevalence.
% AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.

The statistical framework developed for HIV is applicable to SARS-CoV-2.
(see \cref{E-inc-prev:sec:infection-process}).
However, unlike with HIV/AIDS, recovery from SARS-CoV-2 infection is possible, so recoveries need to be considered in the backcalculation.
This introduces a dependency structure in the prevalence data (see \cref{E-inc-prev:sec:prevalence-process}).
The development of this framework makes clear that the duration of positivity (how long an individual tests positive for SARS-CoV-2) is the key quantity relating prevalence to incidence.
Therefore, estimation of duration is central to this thesis (see \cref{intro:sec:duration}).

Furthermore, prevalence surveys are noisy because they are based on population sampling.
The original formulations of backcalculation neglected any observation noise.
AIDS is almost always diagnosed and recorded in developed countries.
That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
An exception is that the reporting delay may be important~\autocite{paganoHIV}; but reporting delay is not important for this thesis's work.
Since its development, backcalculation has been further developed for a wide-variety of contexts.
I incorporate observation noise into the backcalculation approach in \cref{inc-prev:sec:observation-process}.
% Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.

\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
They used a Gaussian process prior to smooth the incidence, applying the prior to either smooth the log-incidence or its derivative; smoothing of the incidence is required to stabalise the estimates~\autocite{brookmeyerBackcalculation}.
\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the backcalculation, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.

\subsection{Mechanistic models}

\subsubsection{Background}

Mechanistic models make assumptions on the biological and epidemiological mechanisms driving transmission~\autocite{lesslerMechanistic}.
Therefore, the model's state and parameters have a biological and/or epidemiological interpretation.
% Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
In particular, this allows changes in transmission to be decomposed into the contributing factors, \eg the number of contacts per day or the probability of transmission upon contact (see \cref{E-SEIR:sec:mechanistic-models}).
The ability to estimate the contribution of each of these factors can inform public health policy.
Furthermore, mechanistic models can be used for scenario modelling, which simulates the effect of interventions.
A scenario is the effect of a proposed intervention (or lack thereof), simulated by modifying the relevant parameters.
% The mechanistic model can then be simulated forward in time to understand the intervention's effects.
% All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
% However, they make stronger assumptions about the disease and population behaviour than statistical models.

The literature on mechanistic models is vast, having been developed for over a century.
The model I introduce in \cref{E-SEIR:sec:SIR} was first formulated by \textcite{kermackContribution}.
However, the earliest mechanistic model to be formulated mathematically was probably \textcite{rossMalariaA}'s model of malaria transmission~\autocite{lesslerMechanistic}.
This early contribution led to an important insight for public health, showing that malaria could be controlled even if mosquitoes were not fully eliminated.

\Textcite{kermackContribution} was a seminal paper for the field, explaining the observation that epidemics ended before all individuals were infected.
They developed a model showing that a population-level build-up of immunity known as \emph{herd immunity} is sufficient to explain this observation.
Their model has inspired much of the subsequent literature, a full review of which is beyond the scope of this thesis.
The relevant background information is explained in \cref{E-SEIR:sec:mechanistic-models}.

Two types of mechanistic models were prominent during the SARS-CoV-2 pandemic: those based on a \emph{renewal equation} and those with a \emph{compartmental} structure.

The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
\begin{align}
    I(t) &= \int_{0}^{t} \beta(t,\tau) I(t-\tau) d \tau
\end{align}
where $I(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature in this thesis, for further information, including practical considerations when using them as the basis of a statistical model, see \textcite{nashEstimating,gosticPractical}.

Compartmental models split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments.
Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
Unlike renewal equations, there are not widely-used implementations of compartmental models.
A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

Compartmental models can be fitted to almost any data source, and commonly incorporate multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.

\subsubsection{Application to SARS-CoV-2}
{\color{red} you are missing a big opportunity here. You could elaborate a bit more the SPIMO part to explain what kind of data people used and introduce the question of whether the same evidence could be provided just by using data from a CIS type model. This is what motivates your research. Other people have used (only?) prevalence data (you give a lot of references below)..but have they recosntructed incidence and transmission only using prevalence data?}

Many studies involving mechanistic models were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), using this type of models to generate evidence for policy development~\autocite{medleySPIM,govSPIMO}.
For example, each of the stages of lifting the lockdown in 2021 was supported by  modelling studies informing the government's decision~\autocite{sageEvidence}.

Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
The nowcasting estimates were mainly based on mechanistic models.
These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).

In this thesis, I focus on the use of data from prevalence surveys.
Specifically, surveys which collect swabs from individuals for testing by RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}).
Several compartmental models incorporated data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
These models incorporate prevalence survey data by adding additional compartments representing RT-PCR positive individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
The proportion of the population in these compartments are then fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.

\Textcite{nicholsonImproving} have the most sophisticated use of prevalence survey data.
They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:duration-SARS-CoV-2} for discussion of the \textcite{hellewellPCRSensitivity} study).
Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.
% However, they do not incorporate age-stratification into their transmission model, and primarily aim to estimate the bias of community testing in the UK rather than incidence and/or transmission.

\todo[inline]{Maybe the following paragraph should be in the contribution section?}
In \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
I allow a more flexible distribution of the duration of positivity than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:duration-SARS-CoV-2}).
Unlike \textcite{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
I use only data from the CIS, and show this is sufficient for estimating incidence and transmission.
Finally, I fit directly to counts from the CIS, rather than an approximation of this data, which has not previously be done.

\subsection{Other approaches}

Other approaches have been taken to estimate incidence and/or transmission from the CIS data.

    \section{Estimating duration} \label{intro:sec:duration}

The distribution of the duration of positivity is central to the relationship between incidence and prevalence (see \cref{E-inc-prev:sec:prevalence-process}).
Therefore, a reliable estimate of this quantity is required to achieve this thesis's aims.

\subsection{Methodology}

Two approaches are commonly used to estimate the duration of a disease: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing.

The biomarker approach, which I take in \cref{E-ATACCC}, models how the amount of virus in an individual varies over time.
The RT-PCR tests commonly used give a proxy for viral load for positive tests (see \cref{E-biology-data:sec:PCR-process}).
Modelling a biomarker uses more information per observation, but requires stronger assumptions.
For example, a functional form of the behaviour over time of the biomarker is required.
In addition, if individual-level variation wants to be included, which is generally important, a distributional form of random effects is required.
This requires both sufficient observations per individual and an assumption on the form of the distribution.

Modelling the duration directly, as I do in \cref{E-perf-test,E-imperf-test} considers only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted.
This approach generally requires fewer parameters, and no random effect(s), reducing the computational burden.
It is also easier to specify non-parameteric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
A more detailed review of the literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.

In both cases, longitudinal data is required.
Therefore, cohort studies, where the same group of indivdiuals are followed over time, are the most appropriate study design.

\subsection{SARS-CoV-2} \label{intro:sec:duration-SARS-CoV-2}

@@

\subsubsection{Previous estimates}

@@

\subsection{Markov chain Monte Carlo} \label{intro:sec:MCMC}

MCMC is a class of algorithms for sampling from a distribution where the normalising constant for the distribution is unknown.
I will use it to sample a set of parameters, say $\psi$ (which may be a scalar or vector), from a posterior distribution $p(\psi \mid y)$ where $y$ is the data (again, scalar or vector).
MCMC algorithms produce a sample $\psi^1, \psi^2, \dots, \psi^M$ from $p(\psi \mid y)$ by constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution~\autocite[275]{gelmanBDA}.
It can be proved that as $M\to\infty$, the sample tends towards a sample from the stationary distribution, \ie the target posterior.
Further details of MCMC are given in \cref{E-MCMC}.
Two practical issues arise when applying MCMC: \emph{convergence} and \emph{stability}~\autocite[72]{lunnBUGS}.

Convergence is how close a MCMC chain's samples are to its stationary distribution.
Early draws from the Markov chain reflect the choice of the starting point, $\psi^0$, rather than the stationary distribution that the chain will eventually converge to~\autocite[282]{gelmanBDA}.
Various metrics quantify this, I will make use of the improved Rhat statistic~\autocite{vehtariRhat}.
%The Rhat statistic is defined for a scalar parameter $\psi$.
%Let $\psi^{(ij)}$ denote the $i$th (of $M$) sample from the $j$th chain (of $N$).
%Let $\bar{\psi}^j$ denote the mean value of the samples of $\psi$ from the $j$th chain, $\bar{\psi}^j = \frac{1}{M} \sum_{i=1}^M \psi^{(ij)}$.
%Let $\bar{\psi}$ denote the mean value of the samples of $\psi$ from all chains, $\bar{\psi} = \frac{1}{N} \sum_{j=1}^N \bar{\psi}^j$.
%Then the between-chain variance is $B = \frac{M}{N-1} \sum_{j=1}^N (\bar{\psi}^j - \bar{\psi})^2$.
%The within-chain variance is $W = \frac{1}{N} \sum_{j=1}^N \frac{1}{M-1} \sum_{i=1}^M (\psi^{(ij)} - \bar{\psi}^j)^2$.
A Rhat of 1 means that the within-chain and between-chain variance are equal, informally that convergence is perfect, as far as can be measured by this technique.
Various rules-of-thumb exist for what should be considered an acceptable Rhat.
Historically, 1.1 has commonly been recommended, although more recent work suggests that 1.01 or 1.05 is more appropriate~\autocite{vehtariRhat}.

The stability of estimates considers the error arising from only using a finite sample of the distribution and the samples being autocorrelated.
Autocorrelation means that estimates using the sample have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
The \emph{effective sample size} corrects for the autocorrelation.
The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocites[286]{gelmanBDA}{vehtariRhat}.
Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}, implemented in RStan~\autocite{RStan-2-32-3}.

% \section{Notational conventions}

% \begin{itemize}
%     \item Lower-case variables are realisations of the respective upper-case random variables.
%     \item Matrices and vectors are bold and have elements with the same unbolded letter, \eg $\vec{x} = [x_1, x_2, \dots]^T$
%     \item $\cdot$ denotes element-wise multiplication
%     \item Distributions used defined in \cref{E-distributions}
% \end{itemize}

@@

First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
A deterministic backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
This approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
I show that, this is a good approximation as long as the proportion of the population sampled is small.
In particular, it does not require the prevalence or incidence to be large.

@@

Finally, in \cref{E-SEIR}, I estimate the transmission of SARS-CoV-2 in England using a mechanistic model, from only the CIS data.
This includes the development of a novel observation model to link the mechanistic transmission model to the data.
The observation model allows the inclusion of the skew in the duration of positivity, without being computationally infeasible.

>>>>

The most famous reproduction number is the \emph{basic reproduction number}, $\R$.
$\R$ is the mean number of secondary infections generated by a single infection in a \emph{fully susceptible} population with \emph{standard behaviour patterns}.
A fully susceptible population means none of its members have any immunity to the disease.
Note that this definition implies $\R$ is constant for a given pathogen and population.
In the early stages of an epidemic caused by a novel pathogen, the number of individuals with immunity is negligible, and hence $\R$ describes this part of the epidemic well.
$\R$ has a threshold value of 1: if $\R < 1$ then the epidemic will die out.

@@

\emph{Prevalence} is the proportion of the population with the disease at a given time.
Whether an individual is \emph{prevalent} (has the disease) is normally measurable.
Therefore, the prevalence of the population can be estimated with a prevalence survey.
However, these are very expensive to power appropriately; SARS-CoV-2 is unusual in that its impact is so large that the UK government did this.
\cref{E-inc-prev:sec:definitions} provides a formal definition of incidence and prevalence.
Prevalence and incidence are strongly related, with the relationship determined by by the duration of an infection (see \cref{E-inc-prev}).

@@

However, incidence and transmission and prevalence, are almost never directly observed.
Therefore, they need to be estimated from other data sources.

Introduce next section eg incidence / transmission not observed directly, need infer from data such as next section.

\section{Data sources}

The data sources used for outbreak surveillance are often observational in nature.
Therefore, they suffer from biases that can be hard to understand and quantify.
For example, community testing which relies on individuals presenting for a test, hence biased by the willingness of individuals to be tested.

Well-designed studies can provide direct estimates of incidence and/or prevalence in specific cohorts.
However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
Furthermore, for these studies to be appropriately powered, they often require a large cohort, and hence are expensive to run.
A unique feature of the COVID-19 pandemic is that its impact was so large that the UK Government were willing to fund such studies.
The availability of this data motivates the work in this thesis.

\section{Transmission model}

The first component required to estimate incidence and transmission is a \emph{transmission model}, a mathematical description of how incidence arises.
A spectrum of models exist, categorised by the strenght of assumptions they make about the transmission process.
Stronger assumptions allow more insight into the transmission process and generally lower the variance of the estimates, but can introduce bias and are often computationally challenging to fit.

The models with the weakest assumptions are statistical models.
These make almost no assumptions about the transmission process, viewing incidence as a generic counting process.
The most assumption is that incidence follows a Poisson process (defined in \cref{E-inc-prev:sec:poisson-process}), and the intensity of the process is the target of incidence.
% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.

Next, are semi-mechanistic models.
These have components of the transmission process in them but do not fully represent the process.
The most common type of semi-mechanistic model are those based on \emph{renewal equation}.
The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
\begin{align}
    I(t) &= \int_{0}^{t} \beta(t,\tau) I(t-\tau) d \tau
\end{align}
where $I(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
The effective reproduction number can be constructed as a function of $\beta(t, \tau)$.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature futher in this thesis, for details, including practical considerations when using them as the basis of a statistical model, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex and aim to fully represent the transmission process.
\emph{Compartmental models} are the most common type of mechanistic model.
Compartmental models split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments.
Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
Unlike renewal equations, there are not widely-used implementations of compartmental models.
A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

\section{Observation models}

The \emph{observation model} links the transmission model to the data.
The choice of observation model can, in theory, be independent of the transmission model.
However, pragmatic considerations often mean that the observation model is chosen to be in a similar form to the transmission model.

Typically, the core of an observation model is a \emph{convolution equation} of an approximation of it.
This equation links an observation at time $t$, $y_t$, to the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    y_t &= \zeta \sum_{i=0}^\infty I(t-i) f(i)
\end{align}
where $I(t)$ is the incidence at time $t$; $\zeta$ is the proportion of infections that are ever observed; and $f(i)$ is the probability that an infection at time $t-i$ is observed at time $t$, conditional on it ever being observed.

% The choice of an observation model must be heavily informed by the data that will be used.
% Two types of data commonly arise in the context of infectious disease: incidence measures and prevalence measures.

% An incidence measure is characterised by each infection only appearing in the dataset at most once.
% Common examples are hospital admissions and deaths.
% The basic form of an observation model for an incidence measure is a scaling and delay of the incidence.
% The scaling is the proportion of infections that are observed (\eg the proportion of infections that are fatal), and the delay is the time between an infection occurring and it being observed (\eg the time between infection and death).
% In discrete time this is:
% \begin{align}
%     M(t) = \zeta \sum_{\tau=0}^\infty I(t-tau) f(\tau)
% \end{align}
% where $M(t)$ is the observation of the incidence measure at time $t$; $I(t)$ is the incidence at time $t$; $\zeta$ is the scaling; and $f(\tau)$ is the delay, formalised as the probability that an infection is observed $\tau$ days after it occurs, given that it will be observed.

% A prevalence measure is characterised by each infection possibly appearing in the dataset multiple times.
% Common examples are the results of a prevalence survey or the number of individuals in hospital with a disease.
% \todo[inline]{Finish writing about prevalence measures}

Solving a convolution equation for $I$ is an ill-posed problem when $f(0)$ is small.
Either an approach which gives a large $f(0)$, for example by assuming that the delay between incidence and being prevelant is 0, or smoothing on $I(t)$ is required.

For a prevalence survey, the delay distribution is the \emph{duration of positivity}, the distribution of the time between an individual becoming infected and the time they are no longer infected.

\section{Backcalculation}

\emph{Backcalculation} was a modelling framework developed in the context of HIV/AIDS but since applied more broadly.
The aim of backcalculation is to estimate the incidence of a disease from the observation of a prevalence measure.
It combines a statistical transmission model with

\section{Estimating the duration of positivity}

Two approaches are commonly used to estimate the duration of a disease: either modelling an underlying biomarker or modelling the duration directly~\autocite[e.g.][]{sweetingEstimating}.
Modelling a biomarker uses more information per observation, but requires stronger assumptions.

The biomarker approach models the trajectory of a biomarker over time in a cohort of individuals.
An individual is considered to have the disease while the biomarker is above or below some threshold.
Parametric assumptions are commonly required for this approach.
For example, a functional form of the behaviour over time of the biomarker is required.
In addition, if individual-level variation wants to be included, which is generally important, a distributional form of random effects is required.
This requires both sufficient observations per individual and an assumption on the form of the distribution.

Modelling the duration directly considers only the binary result at each test (positive or negative); for instance in a \emph{survival analysis} framework.
This approach generally requires fewer parameters, and no random effect(s), reducing the computational burden.
It is also easier to specify non-parameteric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
% A more detailed review of the literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
% Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.

In both cases, longitudinal data is required.
Therefore, cohort studies, where the same group of individuals are followed over time, are the most appropriate study design.

\section{Application to SARS-CoV-2 in England}

The application I will consider in this thesis is the SARS-CoV-2 epidemic in England.

\subsection{Data sources}

\subsubsection{For estimating incidence and transmission}

The main data sources informing the response to the SARS-Cov-2 epidemic in England were: community testing (known as pillar 2 testing), the number of severe events (\eg hospital admissions or deaths), seroprevalence surveys (estimating the proportion of the population with antibodies to a disease, a marker of prior infection), and prevalence surveys (estimating the proportion of the population currently with the disease).
Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} for further details.

For large parts of the pandemic, community testing was widely available in the UK.
The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
However, there were many test results, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.

Seroprevalence surveys indicate the proportion of the population that have been infected with the disease.
Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.
Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) who are not representative of the general population.

Severe events counts are generally a reliable measure of the burden of the disease.
The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context\todo{ref severe events criteria}.
Therefore, these data are subject to fewer biases than community testing.
However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe event counts are insensitive to changes in transmission only occurring in younger individuals.

@@

The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
These surveys use RT-PCT (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal study, testing a large number of indidividuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
Early in the pandemic, a methodology was developed for using CIS data to estimate prevalence (see \cref{E-intro:sec:cis}).

@@

@@

\subsection{SPI-M-O}

In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), using this type of models to generate evidence for policy development~\autocite{medleySPIM,govSPIMO}.
For example, each of the stages of lifting the lockdown in 2021 was supported by  modelling studies informing the government's decision~\autocite{sageEvidence}.

Each week, the members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
The consensus value was published on the UK Government's website and informed its response to the epidemic~\autocite{govRnumber}.

The nowcasting estimates were mainly based on mechanistic and semi-mechanistic models with a wide variety of data sources.
The mechanistic models commonly incorporated multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
The models used were reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).

I will investigate whether estimates relying on only one data source, the CIS, could replace this process.
Such estimates should be more robust, due to the representative nature of CIS.

\subsection{Use of prevalence surveys}

Several previous studies have used SARS-CoV-2 prevalence surveys to estimating incidence and/or transmission.

The \emph{backcalculation} framework~\autocite[e.g.][]{brookmeyerMeasuring,brookmeyerMethod,brookmeyerBackcalculation}, developed in the context of HIV/AIDS
\todo{this is a bad place for an intro to backcalculation: where does it go in this structure?}

\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
They used a Gaussian process prior to smooth the incidence, applying the prior to either smooth the log-incidence or its derivative; smoothing of the incidence is required to stabilise the estimates~\autocite{brookmeyerBackcalculation}.
\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the backcalculation, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.

Several compartmental models incorporated data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
These models incorporate prevalence survey data by adding additional compartments representing RT-PCR positive individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
The proportion of the population in these compartments are then fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.

\Textcite{nicholsonImproving} have the most sophisticated use of prevalence survey data.
They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:duration-SARS-CoV-2} for discussion of the \textcite{hellewellPCRSensitivity} study).
Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.

\subsection{Previous duration estimates}

@@

@@

First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
A \emph{deterministic} backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
The deterministic approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
I show that, this is a good approximation as long as the proportion of the population sampled is small.
In particular, it does not require the prevalence or incidence to be large.
The estimates I produce in \cref{E-backcalc} fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of positivity from a general population study, unlike prior work.

@@

Finally, in \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
I allow a more flexible distribution of the duration of positivity than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:duration-SARS-CoV-2}).
Unlike \textcite{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
I use only data from the CIS, and show this is sufficient for estimating incidence and transmission.
